---
title: "Term diposits campaign"
author: "Jose M Sallan"
date: "22 de octubre de 2019"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective of the analysis

We want to answer the following question:

Which kind of customers are most likely to subscribe (yes/no) a term deposit (variable y)?

This means that we need to build a model to predict if a customer will suscribe a term deposit.

# Methdology

## The dataset

The variables in our dataset are:

1. age (numeric)
2. job: type of job (categorical)
3. marital: marital status (categorical)
4. education: educational level (categorical)
5. default: has credit in default? (binary)
6. balance: average yearly balance, in euros (numeric) 
7. housing: has housing loan? (binary)
8. loan: has personal loan? (binary)
From here, information related with the last contact of the current campaign:
9. contact: contact communication type (categorical) 
10. day: last contact day of the month (numeric)
11. month: last contact month of year (categorical)
12. duration: last contact duration, in seconds (numeric)
Other attributes to consider:
13. campaign: number of contacts performed during this campaign and for this client (numeric)
14. pdays: number of days that passed by after the client was last contacted from a previous campaign (numeric, -1 means client was not previously contacted)
15. previous: number of contacts performed before this campaign and for this client (numeric)
16. poutcome: outcome of the previous marketing campaign (categorical)
Output variable (desired target):
17. y: has the client subscribed a term deposit? (binary)

We can classify the variables as follows:

* Demographics (variables 1 to 4)
* Description of each individual as a customer (variables 5 to 8)
* Impact of campaign on customer (variables 9 to 14)
* Impact of previous campaigns on customer (variables 14 a 16)
* Target variable (variable 17)

# Results

## Examining data

We will start loading the data into R being categorical variables as factors, and summarise the resulting data frame:

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
banks01 <- read.csv2("bank.csv")
summary(banks01)
```

The summary allows us to see that there are no missing data in any variable, and the categories available for each variable. As for the outcome variable we learn that `r sum(banks01$y=="yes")` have requested to open a term diposit, and `r sum(banks01$y=="yes")`, so that a `r sum(banks01$y=="yes")*100/nrow(banks01)`% of customer have suscribed a bank loan.

```{r}
banks01 %>% ggplot(aes(x=y, y=age, fill=y)) + geom_boxplot()
```


```{r}
banks01 %>% group_by(job) %>% summarise(yes=sum(y=="yes")/n(), no=sum(y=="no")/n()) %>% gather(value, proportion, -job) %>% ggplot(aes(x=job, y=proportion, fill=value)) + geom_bar(stat = "identity") + theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
banks01 %>% group_by(marital) %>% summarise(yes=sum(y=="yes")/n(), no=sum(y=="no")/n()) %>% gather(value, proportion, -marital) %>% ggplot(aes(x=marital, y=proportion, fill=value)) + geom_bar(stat = "identity") + theme_bw()
```

```{r}
banks01 %>% group_by(education) %>% summarise(yes=sum(y=="yes")/n(), no=sum(y=="no")/n()) %>% gather(value, proportion, -education) %>% ggplot(aes(x=education, y=proportion, fill=value)) + geom_bar(stat = "identity") + theme_bw()
```



