---
title: "Examining flights from NYC airports with ggplot2"
author: "Jose M Sallan"
date: "04/04/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

ggplot2 is a system for creating graphics based on The *Grammar of Graphics*: 

- Provide the *data*.
- tell `ggplot` how to map variables to aesthetics `aes`.
- tell what *graphical primitives* to use.

To illustrate how can we create plots with `ggplot`, we will use the `nycflights13` data. We will also load the `tidyverse` package:

```{r, message=FALSE}
library(tidyverse)
library(nycflights13)
```

# Preparing `flights` data

We will analyze depart and arrival delays frow airports in New York City in 2013. That data is in the `flights` table of `nycflights13`:

```{r}
flights
```

We can be interested obtaining the following variables for every hour (that is, for every unique `year`, `month`, `day` and `hour`) and every `origin` (as we have three diferent origin airports in this database):

- Average departure delay `av_dep`.
- Average arrival delay `av_arr`.
- The number of flights `traff`.

We can do that using the dplyr verbs for data wranglig:

```{r}
av_delays <- flights %>% filter(!is.na(dep_delay) & !is.na(arr_delay)) %>% group_by(year, month, day, hour, origin) %>% summarise(av_dep=mean(dep_delay), av_arr=mean(arr_delay), traffic=n())
```

To obtain a `ggplot` graph we need:

- to pass the data and map the aesthetics with the `ggplot` function
- to tell the primitives to use.

As we will see, each primitive is associated to an aesthetic

We can select a color by a number or a name from `colors()` ([color chart](http://research.stowers.org/mcm/efg/R/Color/Chart/ColorChart.pdf)) or using other codes like HEX.

We can get different types of graphs for several mappings:

* one variable, categorical or continuous
* discrete variable and continuous variable 
* two continuous variables

We can also create facet grids on a categorical variable.

# One variable

## Categorical variable: barplots with `geom_bar`

We can use **barplots** to visualize the distribution of categorical variables, counting how many elements are in each category. For instance, we can count how many flights are flown by each carrier:

```{r, fig.align="center"}
ggplot(flights, aes(carrier)) + geom_bar()
```

Perhaps we can be interested in knowing how much flights sends each airline from each `origin`: now we are using two categories: carrier and airport. We put a `color` in the aesthetic equal to the `origin` variable.

```{r, fig.align="center"}
ggplot(flights, aes(carrier, fill=origin)) + geom_bar()
```

We can have different presentations using the `position` variable in `geom_bar`.

```{r, fig.align="center"}
ggplot(flights, aes(carrier, fill=origin)) + geom_bar(position = "dodge")
ggplot(flights, aes(carrier, fill=origin)) + geom_bar(position = "fill")
```

## Continuous variables: histograms with `geom_histogram()`

We use histograms to examine the distribution of a continuous variable. To do so, we split the variable into several ranges or bins, and count how many observations fall in each bin. In the histogram, each bin is represented as a bar with a height proportional to the number of observations falling in it. Here we are plotting an histogram of the distribution of the `av_dep` variable.

```{r}
ggplot(av_delays, aes(av_dep)) + geom_histogram()
```

We can improve the appearance of the histogram by modifying the number of bins:

```{r}
ggplot(av_delays, aes(av_dep)) + geom_histogram(bins=10)
```


## One-dimension density plots: `geom_density()`

A `density` is an histogram whose bins are of infinitesimal width. Note the similarity of this plot of the distribution of `av_dep` with the previous histogram.

```{r, fig.align="center"}
ggplot(av_delays, aes(av_dep)) + geom_density()
```

## Modifying graphical parameters

We can modify the graphical parameters, with a lines in blue and density fill in green:
```{r, fig.align="center"}
ggplot(av_delays, aes(av_arr)) + geom_density(color = "blue", fill = "green")
```

We now also change the `theme`:

```{r, fig.align="center"}
ggplot(av_delays, aes(av_arr)) + geom_density(color = "blue", fill = "green") + theme_bw()
```

With a larger font size:
```{r, fig.align="center"}
ggplot(av_delays, aes(av_arr)) + geom_density(color = "blue", fill = "green") + theme_bw(base_size = 14)
```

# Categorical variable and continuous variable

We can evaluate a continuous variable, say `av_arr`, as a function of a categorical variable, the `origin` airport.

## Barplots: `geom_bar()`

In the previous sections, we have used barplots to count how many elements fall within a categorical variable. We can use `geom_bar` with a esthetic including a categorical variable (`origin`) and a integer variable (`traffic`) to sum up the traffic in each airport:

```{r, fig.align="center"}
ggplot(av_delays, aes(origin, traffic)) + geom_bar(stat = "identity")
```

In this case we don't need to count how many observations are in the origin, as the value of the bar is provided by the traffic. That is why we are using `stat = "identity"` with `geom_bar`.

## Boxplots: `geom_boxplot()`

A **boxplot** shows the data distribution of a variable through its *position statistics*:

- first quartile **Q1**: value of the lowest 25% observation of the variable.
- second quartile **Q2** or **median**: value of the observation lying in the 50% of the variable.
- third quartile **Q3**: value of the lowest 75% observation of the variable.

Here is a boxplot for the `av_arr` variable for each category:

```{r, fig.align="center"}
ggplot(av_delays, aes(origin, av_arr)) + geom_boxplot()
```

Let's control the `y limit` so we can appreciate the plot:
```{r, fig.align="center"}
ggplot(av_delays, aes(origin, av_arr)) + geom_boxplot() +
  ylim(c(min(av_delays$av_arr), 100))
```

## Violin plots: `geom_violin()`

A `violin plot` presents a symmetrical picture of the distribution of the variable for each category, similar to the one that we can obtain with `geom_density()`:

```{r, fig.align="center"}
ggplot(av_delays, aes(origin, av_arr)) + geom_violin()
```

Adding `color` and changing the `theme`:

```{r, fig.align="center"}
ggplot(av_delays, aes(origin, av_arr, fill=origin)) + geom_violin() + theme_bw()
```

# Two continuous variables

## Scatterplots: `geom_point()`

We can be interested in checking the relationship between arrival and departure delays for each hour segment, so every observation is a point in a scatterplot. We can also let ggplot draw a tendency line with `geom_smooth`:
```{r, fig.align="center"}
ggplot(av_delays, aes(av_dep, av_arr)) + geom_point() + geom_smooth() + theme_bw()
```

And we can do the same for each origin, plotting it in a different color:
```{r, fig.align="center"}
ggplot(av_delays, aes(av_dep, av_arr, color=origin)) + geom_point() + geom_smooth() + theme_bw()
```

We can also set several labels, titles and captions with `lab`:
```{r}
ggplot(av_delays, aes(av_dep, av_arr, color=origin)) + geom_point() + geom_smooth() + theme_bw() + labs(title="departure vs arrival delays", subtitle="NYC airports", colour="airports", caption="data from nycflights13", x="departure delays", y="arrival delays")
```

## Continuous functions: `geom_line()`

We can also plot the evolution of a function using a set of points using `geom_line()`. First we obtain a `time variable in R time format:

```{r, warning=FALSE}
av_delays <- av_delays %>% mutate( time =  as.Date(paste0(year, "-", month, "-", day, " ", hour), format = "%Y-%m-%d %H", tz = "GMT"))
```

Then we can plot the evolution of arrival delays with time:
```{r, align="center"}
ggplot(av_delays, aes(time, av_arr)) + geom_line()
```

## Two-density plots: `geom_density_2d()`

Similarly to one-density plots, we can draw density plots of bivariates:

```{r, fig.align="center"}
ggplot(av_delays, aes(av_dep, av_arr)) + geom_density_2d() + theme_bw()
```

# Facet grids: `facet_grid()`

When we make plots with categories, we might want to get one plot for each category instead of a plot for each category. We can achieve this with `facet_grid`. We can make grids by rows, by columns, or both. Let's see it with several scatterplots:
```{r, fig.align="center"}
ggplot(av_delays, aes(av_dep, av_arr)) + geom_point() + geom_smooth() + theme_bw() + facet_grid(origin ~ .)
ggplot(av_delays, aes(av_dep, av_arr)) + geom_point() + geom_smooth() + theme_bw() + facet_grid(. ~ origin)
```
