---
title: "Some data.table usage on nycflights13"
author: "Jose M Sallan"
date: "08/04/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a small vignette showing some functionalities of data.table. The original reference of the functionalities shown here can be found at:

* Introduction: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
* Reference semantics: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html



# Loading packages and preparing data

We will use `nycflights13` to pick data, and `data.table` to wrangle data.

```{r}
library(nycflights13)
library(data.table)
```

We have to convert data frames into the `data.table` format:

```{r}
dt_flights <- data.table(flights)
dt_airlines <- data.table(airlines)
dt_airports <- data.table(airports)
```


# Obtaining delays

We may want to turn to zero negative delays. We can use two reference semantics syntaxes to do that. The first is:

```{r, eval=FALSE}
dt_flights[, c("arr_delay", "dep_delay"):=.(ifelse(arr_delay > 0, arr_delay, 0), ifelse(dep_delay > 0, dep_delay, 0))]
```

and the second (more practical, in my opinion):

```{r}
dt_flights[, `:=`(arr_delay=ifelse(arr_delay > 0, arr_delay, 0), dep_delay=ifelse(dep_delay > 0, dep_delay, 0))]
```

Let's look at `dt_flights`:

```{r}
dt_flights
```

# Obtaining a specific flight

What was the fourth flight to depart from each origin airport on 13 February?

```{r}
dt_flights[order(time_hour)][month==2 & day==13, .SD[4], origin]
```

And the last flight?

```{r}
dt_flights[order(time_hour)][month==2 & day==13, .SD[.N], origin]
```

We have use the `.SD` syntax to subset the table internally for each group (see introductory vignette).

# Average delay per destination airport

Let's compute the average and departure delays for each destination airport:

```{r}
dt_flights[, .(mean(dep_delay), mean(arr_delay)), dest]
```

This is ugly. Let's consider that we need to remove not available observations and put a name for each variable:

```{r}
delay_dest <- dt_flights[, .(av_dep=mean(dep_delay, na.rm = TRUE), av_arr=mean(arr_delay, na.rm = TRUE)), dest]
```

Now this is much nicer:

```{r}
delay_dest
```


# What is the name of the destination airport with maximum arrival delay?

To get the name of airports we need `dt_airports`:

```{r}
dt_airports
```

Let's bind `delay_dest` and the columns we need from `dt_airports` using `dest` and `faa` as binding columns:

```{r}
delay_dest <- merge(delay_dest, dt_airports[, .(faa, name)], by.x="dest", by.y="faa", all.x=TRUE)
```

Now we have the name:

```{r}
delay_dest
```

Let's order the destination airports by delay:

```{r}
delay_dest[,.(name, av_arr)][order(-av_arr)]
```

And pick the name of the first destination airport:

```{r}
delay_dest[,.(name, av_arr)][order(-av_arr)][1, name]
```

# What is the name of the airline with maximum arrival delay?

We need to do the same thing for `carrier` as we have done with airports:

```{r}
delay_airline <- dt_flights[, .(av_dep=mean(dep_delay, na.rm = TRUE), av_arr=mean(arr_delay, na.rm = TRUE)), carrier]
```

Carrier names are in `dt_airlines`:

```{r}
dt_airlines
```

As the merging variables are the same in both data frames, we can override `by.x` and `by.y` here:

```{r}
delay_airline <- merge(delay_airline, dt_airlines, all = TRUE)
```

And select the airline with maximum arrival delay:

```{r}
delay_airline[,.(name, av_arr)][order(-av_arr)]
delay_airline[,.(name, av_arr)][order(-av_arr)][1, name]
```

