---
title: "Wrangling data with dplyr"
author: "Jose M Sallan"
date: "29/9/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# The mtcars data

The `mtcars` data was extracted from the 1974 *Motor Trend US magazine*, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models). The meaning of each variable is:

* **mpg**	Miles/(US) gallon
* **cyl**	Number of cylinders
* **disp**	Displacement (cu.in.)
* **hp**	Gross horsepower
* **drat**	Rear axle ratio
* **wt**	Weight (1000 lbs)
* **qsec**	1/4 mile time
* **vs**	Engine (0 = V-shaped, 1 = straight)
* **am**	Transmission (0 = automatic, 1 = manual)
* **gear**	Number of forward gears
* **carb**	Number of carburetors

Usually we are interested in finding predictors of fuel consumptio, measured with the `mpg` variabnle.

```{r}
data("mtcars")
head(mtcars)
```

```{r}
str(mtcars)
```

# Data wrangling

In this document we will use the `dplyr` functions with the `mtcars` data frame.

```{r, message=FALSE}
library(dplyr)
```

## `select` columns

Let's how do we `select` columns with dplyr:

```{r}
head(select(mtcars, mpg, wt, am))
```

We can pick columns with names starting or ending like a specific string:

```{r}
head(select(mtcars, starts_with("c")))
head(select(mtcars, ends_with("p")))
```

... and we can store the selected columns in another data frame:

```{r}
MilesHp <- select(mtcars, c(mpg, hp))
head(MilesHp)
```

## `filter` rows

We can `filter` cases (rows) according to a logical expression. There we are selecting cases with `mpg` lower than 16:

```{r}
filter(mtcars, mpg < 16)
```

We can `filter` using more than one variable:

```{r}
filter(mtcars, mpg < 16 & am==1)
```

## `arrange` (order) by variables

We can `arrange` a data frame ordering it with the values of a column. By default, rows are order in increasing order (from smaller to bigger):

```{r}
head(arrange(mtcars, mpg), 3)
tail(arrange(mtcars, mpg), 3)
```

We set the order using two or more variables, and use `desc` to order in decreasing order (from bigger to smaller):

```{r}
head(arrange(mtcars, am, desc(mpg)), 4)
tail(arrange(mtcars, am, desc(mpg)), 4)
```

Applying a minus sing to de variable makes the same effect as `desc`:

```{r}
head(arrange(mtcars, am, -mpg), 4)
tail(arrange(mtcars, am, -mpg), 4)
```


## `mutate` (create) and `rename` variables

We `mutate` a data frame by adding one or more new columns. To make the addition effective, we must save the results of the addition into a variable, that can be the same data frame:

```{r}
mtcars2 <- mutate(mtcars, l100km = 235.215/mpg, model=rownames(mtcars))
head(mtcars2)
```

We can also `rename` existing columns using the convention *new name = old name*:

```{r}
mtcars2 <- rename(mtcars2, manual = am)
head(mtcars2)
```

## The piping operator

It is frequent that we need to use several dplyr verbs to obtain some result. For instance, we may want to filter to some subset of elements, and select only some variables. This may require making several intermediate steps:

```{r}
aux1 <- filter(mtcars2, manual==1 & vs==1)
result <- select(aux1, model, l100km, wt, disp)
result
```

We can perform several operations with a single instruction using the **piping operator** `%>%`. When using the piping operator, we omit the first argument of the dplyr functions -the data frame-, that is replaced by the operations performed at the left of each operator:

```{r}
result2 <- mtcars2 %>% filter(manual==1 & vs==1) %>% select(model, l100km, wt, disp)
result2
```

The result of both operations is identical:

```{r}
identical(result, result2)
```


# Grouping

## Defining groups

When whe make an operation in a data frame, it is applied to the whole data frame by default. It is frequent to include categorical variables in the data, that allow grouping observations (e.g., automatic or manual, or by number of gears) using the `group_by` function. When we apply `group_by` to a data frame, nothing happens apparently, but the data frame has been sliced into as many groups as levels of the categorical variable.

Let's apply that grouping functionality to define an `id` number for cars with automatic and manual transmission combining `group_by` and `mutate`:

```{r}
mtcars2 <- mtcars2 %>% group_by(manual) %>% mutate(id=1:n())
```

Let's see how `id` has been defined for each group:

```{r}
mtcars2 %>% filter(manual == 0) %>% select(manual, model, id) %>% print(n=5)
mtcars2 %>% filter(manual == 1) %>% select(manual, model, id) %>% print(n=5)
```


## Summarising and grouping

We can `summarise` a data frame obtaining some statistic values for their columns. The result is a single row with the values for each column. In this case, we find the average of consumption and weight:

```{r}
mtcars2 %>% summarise(cons=mean(l100km), weight=mean(wt))
```

Using `group_by` together with `summarise` we can summarise values for each category. Here we are obtaining mean values of consumption and weight for automatic and manual cars:

```{r}
mtcars2 %>% group_by(manual) %>% summarise(cons=mean(l100km), weight=mean(wt))
```

We can `group_by` more than one category. Then, we obtain the summarised values for all combinations of categories existing in the data set:

```{r}
mtcars2 %>% group_by(gear, manual) %>% summarise(cons=mean(l100km), weight=mean(wt))
```

In the above example, we see that there are no rows with `gear=5` and `manual=0`, and with `gear=3` and `manual=1.



# Joining dataframes

dplyr includes a set of functions for joining dataframes with a relational structure. This means that we have columns in each table (not necessarily with the same name), that can relate the contents of the two tables.

```{r, message=FALSE}
library(dplyr)
```

## Mutating joins

**Mutating joins** return a data frame with the columns of two data frames *x* and *y*:

```{r, echo=FALSE}
url <- "http://www1.udel.edu/evelyn/SQL-Class2/joins.jpg"
```

![](`r url`)

* **inner join**: returns all rows from *x* where there are matching values in *y*. If there are multiple matches between *x* and *y*, all combination of the matches are returned.

* **full (outer) join**: returns all rows from both *x* and *y*.  Where there are not matching values, returns NA for the ones missing.

* **left join**: returns all rows from **x**. Rows of *x* with no match in *y* will have NA values in the *y* columns. If there are multiple matches between *x* and *y*, all combination of the matches are returned.

* **right join**: returns all rows from **y**. Rows of *y* with no match in *x* will have NA values in the *x* columns. If there are multiple matches, all combinations are returned.

These four mutating joins are implemented in dplyr with the **inner_join**, **full_join**, **left_join** and **right_join** functions.

To illustrate the result of the four mutating joins, we use a minimal example from the `dplyr`package:

```{r}
band_members
band_instruments
```

The columns used to build the join are, by default, the ones with the same column name, in this case the `name` columns. Let's do a inner and and outer join:

```{r}
inner_join(band_members, band_instruments)
full_join(band_members, band_instruments)
```

And now let's perform a left and right join:

```{r}
right_join(band_members, band_instruments)
left_join(band_members, band_instruments)
```

## Filtering joins

Filtering joins are used to retain (filter) rows of *x* with matching values in *y*, or without matching values in *y*. In dplyr, we use the `semi_join` and `anti_join` functions to do the first and second filtering, respectively:

```{r}
semi_join(band_members, band_instruments)
anti_join(band_members, band_instruments)
```

## Specifying joining columns with by

If the names of joining columns are different in each data frame, we use the parameter `by`:

```{r}
inner_join(band_members, band_instruments2, by=c("name" = "artist"))
```

Note that in the result, the name of the joining column of the left-hand side data frame is retained.










